|

### 2024-02-15 paper/paper.Rmd
update paper

---
title: |
  Automated assessment of residual plots with computer vision models
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
  - name: Klaus Ackermann
    affil: a
    email: Klaus.Ackermann@monash.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  TBD.
keywords: |
  TBD
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

\newpage
\tableofcontents
\newpage

@@

However, as pointed out by @li2023plot, a primary limitation of the lineup protocol lies in its reliance on human judgments. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labor costs and time requirements.

@@

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, focusing on scatterplot diagnostics. @wilkinson2005graph expanded on this work, introducing graph theoretic scagnostics, which encompassed computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate actual residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

@@

```{r false-finding, fig.width = 8/5*4, fig.height = 6/4*4, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity."}
set.seed(10131)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

Another design involves using an image as input. The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very. These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input. The main considerations are the image resolution and the aesthetics of the residual plot. In general, higher resolution provides more information to the model but comes with the trade-off of increased complexity and greater difficulty in training. As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model. This implies that the model can not accept arbitrary images as input but requires the use of the same preprocessing pipeline to convert residuals and fitted values into a standardized-style residual plot.

@@

Distribution $Q$ is derived from the correctly specified model. However, if the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different from $Q$. For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but not included in $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$. Besides, if the $\boldsymbol{\varepsilon}$ follows a non-normal distribution such as a multivariate lognormal distribution, $P$ will usually be skewed and has a long tail.

@@

We tried to train a computer vision model to approximate this distance measure with a residual plot. Let $D$ be the result of equation \ref{eq:kl-0} indicating the extent of the model violations, and our estimator $\hat{D}$ is formulated as

@@

With the approximated distance $\hat{D}$, we will be able to know how different the underlying distribution of the residuals is from a good residual distribution. It also provides information for the strength of the visual signal embedded in the residual plot.

@@

@@

Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$. In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and lognormal distributions, as presented in Figure \ref{fig:different-e}. Each distribution imparts unique characteristics to the residuals. The discrete error distribution introduces discreteness in residuals, while the lognormal distribution typically yields outliers. Uniform error distribution may result in residuals filling the entire space of the residual plot. All of these distributions exhibit visual distinctions from the normal error distribution.

@@

For each generated residual plot, we computed four scagnostics – "Monotonic," "Sparse," "Splines," and "Striped" – using the cassowaryr R package [@mason2022cassowaryr]. These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model. Although other scagnostics are informative, they are currently unavailable due to a fatal bug in the compiled C program of the interp R package [@Albrecht2023interp], which may unpredictably crash the process. For reproducibility, we excluded these scagnostics from the training data.

@@

The model begins with an input layer of shape $n \times h \times w \times 3$, capable of handling $n$ RGB images. This is followed by a grayscale conversion layer utilizing the luma formula under the Rec. 601 standard, which converts the color image to grayscale. Grayscale suffices for our task since data points are plotted in black. We experiment with three combinations of $h$ and $w$: $32 \times 32$, $64 \times 64$, and $128 \times 128$, aiming to achieve sufficiently high image resolution for the problem at hand.

@@

To achieve an optimal deep learning model, hyperparameters like the learning rate often need to be fine-tuned using a tuner. In our study, we utilized the Bayesian optimization tuner from the KerasTuner Python library [@omalley2019kerastuner] for this purpose. A comprehensive list of hyperparameters is provided in Table \ref{tab:hyperparameter}.

@@

- RMSE for the test set
- R^2
- Mean bias deviation to understand the overall bias
- quantile loss to understand how well the model captures the entire distribution of D
- Percentage of predictions within a tolerance interval (like 0.1)

- comparison with human visual inference
- overview of the human subject experiment
- metrics
- agreement rate
- check cases that are aggreed
- check cases that are disaggreed

@@

The figure highlights that most under-predictions occurred when $\hat{D} < 3$, while over-predictions occurred predominantly when $3 < \hat{D} < 6$. To comprehend the reasons behind these deviations, we conducted a meta-analysis on $D - \hat{D}$ for all factors used in the data generating process. A regression tree was constructed using the rpart R package [@Terry2022rpart]. The results of the regression tree are provided in the Appendix.

The meta-analysis revealed that most issues arose from non-linearity problems and the presence of a second predictor in the regression model. When the error distribution had a very small variance, all computer vision models tended to under-predict the distance. Conversely, when the error distribution had a large variance, all models tended to over-predict the distance. Therefore, Additionally, for input images representing null plots, it was expected that the models will over-predict the distance, as explained in Section \ref{null-distribution-of-the-approximated-distance}.

@@

model_pred %>%
  ggplot() +
  geom_hline(yintercept = 0, alpha = 0.5) +
  # geom_point(aes(effect_size, effect_size - vss), alpha = 0.03) +
  annotate("rect", 
           ymin = min(model_pred$effect_size - model_pred$vss),
                ymax = 0,
                xmin = min(model_pred$vss),
                xmax = max(model_pred$vss),
            fill = "#66c2a5",
            alpha = 0.3) +
  geom_point(data = NULL, aes(5, 0, col = "over-prediction"), shape = 15) +
  annotate("rect", 
           ymin = 0,
                ymax = max(model_pred$effect_size - model_pred$vss),
                xmin = min(model_pred$vss),
                xmax = max(model_pred$vss),
            fill = "#fc8d62",
            alpha = 0.3) +
  geom_point(data = NULL, aes(5, 0, col = "under-prediction"), shape = 15) +
  geom_hex(aes(effect_size, effect_size - vss), bins = 25) +
  geom_smooth(aes(vss, effect_size - vss), se = FALSE, col = "#e34a33") +
  facet_grid(type ~ res) +
  coord_fixed() +
  ylab(expression(D - hat(D))) +
  xlab(expression(hat(D))) +
  theme_light() +
  scale_fill_continuous(limits = c(1, 200), trans = "log10", low = "#56B1F7", high = "#132B43") +
  scale_color_manual(values = c("under-prediction" = scales::alpha("#fc8d62", 0.3),
                                "over-prediction" = scales::alpha("#66c2a5", 0.3)),
                     breaks = c("under-prediction", "over-prediction")) +
  guides(col = guide_legend(title = NULL, override.aes = list(size = 8), order = 1))
```

@@

## Comparison with human visual inference

@@

For each lineup used in @li2023plot, there is one actual residual plot and 19 null plots. While the distance $D$ for the actual residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero. We have used our optimized computer vision model to predict $\hat{D}$ for both the actual residual plots and the null plots.

@@

Figure \ref{fig:human-mosaic} and Table \ref{tab:human-table} provide insights into the agreement between decisions made by the computer vision model and visual tests conducted by humans. It can be observed that visual tests are less sensitive than the computer vision model, as they reject almost only when the computer vision model does. The agreement rates between visual tests and the computer vision model are 67.96% and 66.84% for residual plots containing heteroskedasticity and non-linearity patterns, respectively.

When plotting the decision against the distance, as depicted in Figure \ref{fig:power}, we observe that for the computer vision model, when the distance $D > 3$, almost all residual plots will be rejected. However, for visual tests conducted by humans, the threshold is higher at $D > 5$. Visual tests also demonstrate low sensitivity to residual plots with small distances compared to the computer vision model.

@@

Figure \ref{fig:delta} indicates that the weighted detection rate increases as the $\delta$-difference increases, particularly when the $\delta$-difference is greater than zero. A negative $\delta$-difference suggests that there is at least one null plot in the lineup with a stronger visual signal than the actual residual plot. In some instances, the weighted detection rate is close to one, yet the $\delta$-difference is negative. This discrepancy implies that the distance measure, or the approximated distance, may not perfectly reflect actual human behavior.

@@

```{r human-table}
experiment %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  group_by(type) %>%
  summarise(n = format(n()), agree = format(sum(model_reject == reject)), rate = format(mean(model_reject == reject), digits = 4)) %>%
  kableExtra::kable("latex", 
                    booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Violations", "\\#Samples", "\\#Agreements", "Agreement rate"),
                    caption = "Summary of decisions made by the visual test conducted by human and decisions made by the computer vision model.")
```

@@

```{r power, fig.pos = "!h", fig.cap = "Comparison of decisions made by visual tests and the computer vision model. The data points are fitted with logistic regression models with no intercept but an offset equals to $log(0.05/0.95)$."}

max_actual_ss <- max(filter(experiment, !attention_check, !null_lineup)$actual_ss)

model_glm_pred_poly <- glm(model_reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_actual_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_actual_ss, 0.01), type = "Non-linearity", name = "Computer vision model", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

visual_glm_pred_poly <- glm(reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_actual_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_actual_ss, 0.01), type = "Non-linearity", name = "Visual test", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

conv_glm_pred_poly <- glm(conventional_reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_actual_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_actual_ss, 0.01), type = "Non-linearity", name = "conventional", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))

>>>>

---
title: |
  Automated assessment of residual plots with computer vision models
type: Draft paper (Incomplete)
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
  - name: Klaus Ackermann
    affil: a
    email: Klaus.Ackermann@monash.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  TBD.
keywords: |
  TBD
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

<!-- \newpage -->
<!-- \tableofcontents -->
<!-- \newpage -->

@@

However, as pointed out by @li2023plot, a primary limitation of the lineup protocol lies in its reliance on human judgments. Unlike conventional statistical tests that can be performed computationally in statistical software, the lineup protocol requires human evaluation of images. This characteristic makes it less suitable for large-scale applications, given the associated high labour costs and time requirements.

@@

The utilization of computers to interpret data plots has a rich history, with early efforts such as "Scagnostics" by @tukey1985computer, focusing on scatter plot diagnostics. @wilkinson2005graph expanded on this work, introducing graph theoretic scagnostics, which encompassed computable measures applied to planar proximity graphs. These measures, including, but not limited to, "Outlying", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy", and "Striated" aimed to characterize outliers, shape, density, trend, coherence and other characteristics of the data. While this approach has been inspiring, there is a recognition [@buja2009statistical] that it may not capture all the necessary visual features that differentiate actual residual plots from null plots. A more promising alternative entails enabling machines to learn the function for extracting visual features from residual plots. Essentially, this means empowering computers to discern the crucial visual features for residual diagnostics and determining the method to extract them.

@@

```{r false-finding, fig.width = 8/5*4, fig.height = 6/4*4, fig.pos="!h", fig.cap="An example residual vs fitted values plot (red line indicates 0). The vertical spread of the data points varies with the fitted values. This often indicates the existence of heteroskedasticity. The Breusch-Pagan test rejects this residual plot at 95\\% significance level ($p\\text{-value} = 0.046$)."}
set.seed(452)
ori_x <- rand_lognormal()
mod <- heter_model(b = 0, x = closed_form(~-ori_x))
ori_dat <- mod$gen(300)

@@

Another design involves using an image as input. The primary advantage of this design, as opposed to the vector format, is the availability of the existing and sophisticated image processing architectures developed over the years, such as the VGG16 architecture proposed in @simonyan2014very. These architectures can effectively capture and summarize spatial information from nearby pixels, which is less straightforward with vector input. The main considerations are the image resolution and the aesthetics of the residual plot. In general, higher resolution provides more information to the model but comes with the trade-off of increased complexity and greater difficulty in training. As for the aesthetics of the residual plot, a practical solution is to consistently present residual plots in the same style to the model. This implies that the model can not accept arbitrary images as input but requires the use of the same pre-processing pipeline to convert residuals and fitted values into a standardized-style residual plot.

@@

Distribution $Q$ is derived from the correctly specified model. However, if the model is misspecified, then the actual distribution of residuals denoted as $P$, will be different from $Q$. For example, if the data generating process contains variables correlated with any column of $\boldsymbol{X}$ but not included in $\boldsymbol{X}$, causing an omitted variable problem, $P$ will be different from $Q$ because the residual operator obtained from the fitted regression model will not be the same as $\boldsymbol{R}$. Besides, if the $\boldsymbol{\varepsilon}$ follows a non-normal distribution such as a multivariate log-normal distribution, $P$ will usually be skewed and has a long tail.

@@

We tried to train a computer vision model to approximate this distance measure with a residual plot. Let $D$ be the result of equation \ref{eq:kl-0}, and our estimator $\hat{D}$ is formulated as

@@

With the approximated distance $\hat{D}$, we will be able to know how different the underlying distribution of the residuals is from a good residual distribution. The approximated distance $\hat{D}$ can also be used as an index of the model violations. It also provides information for the strength of the visual signal embedded in the residual plot.

@@

### Approximated distance as an Model Violations Index (MVI)

In Section \ref{output-formats}, we discuss the output formats of the computer vision model presented in this paper. One notable advantage highlighted is its interpretability. The approximated distance $\hat{D}$ is an estimator of the difference between the actual residual distribution and the reference residual distribution. This difference primarily arises from deviations in model assumptions. The magnitude of $D$ directly reflects the degree of these deviations, thus making $\hat{D}$ instrumental in forming a model violation index.

Note that since the Kullback-Leibler divergence could be affected by the number of random variables involved in the evaluation. In general, more observations will result in greater distance $D$. This does not mean $\hat{D}$ can not correctly represent the extent of the model violations since more observations in a residual plot usually lead to stronger visual signal strength as the underlying visual patterns will have higher chance to be revealed except for cases like serious overlapping. For the same data generating process, $D$ usually grows logarithmically as the number of observations increases. This is because $D = \text{log}(1 + D_{KL})$, and $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the independence assumption.

Note that the Kullback-Leibler divergence might be influenced by the number of random variables involved in its evaluation. Generally, a larger number of observations will lead to a greater distance $D$. However, this does not imply that $\hat{D}$ fails to accurately represent the extent of model violations. In fact, when examining residual plots with more observations, we often observe a stronger visual signal strength, as the underlying patterns are more likely to be revealed, except in cases of significant overlapping.

For a consistent data generating process, $D$ typically increases logarithmically with the number of observations. This behaviour comes from the relationship $D = \text{log}(1 + D_{KL})$, where $D_{KL} = \sum_{i=1}^{n}D_{KL}^{(i)}$ under the assumption of independence.

Therefore, the Model Violations Index (MVI) can be proposed as

\begin{equation}
\label{eq:mvi}
\text{MVI} = C + \hat{D} - log(n),
\end{equation}

\noindent where $C$ is a large enough constant keeping the result positive.

@@

Non-normality violations arise from specifying a non-normal distribution for $\boldsymbol{\varepsilon}$. In the synthetic data simulation, four distinct error distributions are considered, including discrete, uniform, normal, and log-normal distributions, as presented in Figure \ref{fig:different-e}. Each distribution imparts unique characteristics to the residuals. The discrete error distribution introduces discreteness in residuals, while the log-normal distribution typically yields outliers. Uniform error distribution may result in residuals filling the entire space of the residual plot. All of these distributions exhibit visual distinctions from the normal error distribution.

@@

For each generated residual plot, we computed four scagnostics – "Monotonic," "Sparse," "Splines," and "Striped" – using the `cassowaryr` R package [@mason2022cassowaryr]. These computed measures, along with the number of observations from the fitted model, were provided as the second input for the computer vision model. Although other scagnostics are informative, they are currently unavailable due to a fatal bug in the compiled C program of the `interp` R package [@Albrecht2023interp], which may unpredictably crash the process. For reproducibility, we excluded these scagnostics from the training data.

@@

The model begins with an input layer of shape $n \times h \times w \times 3$, capable of handling $n$ RGB images. This is followed by a grayscale conversion layer utilizing the luma formula under the Rec. 601 standard, which converts the colour image to grayscale. Grayscale suffices for our task since data points are plotted in black. We experiment with three combinations of $h$ and $w$: $32 \times 32$, $64 \times 64$, and $128 \times 128$, aiming to achieve sufficiently high image resolution for the problem at hand.

@@

To achieve an optimal deep learning model, hyperparameters like the learning rate often need to be fine-tuned using a tuner. In our study, we utilized the Bayesian optimization tuner from the `KerasTuner` Python library [@omalley2019kerastuner] for this purpose. A comprehensive list of hyperparameters is provided in Table \ref{tab:hyperparameter}.

@@

This section remains incomplete as we may modify the "Results" section later on.

@@

The figure highlights that most under-predictions occurred when $\hat{D} < 3$, while over-predictions occurred predominantly when $3 < \hat{D} < 6$. To comprehend the reasons behind these deviations, we fitted a regression tree using the `rpart` R package [@Terry2022rpart] on $D - \hat{D}$ for all factors used in the data generating process. The results of the regression tree are provided in the Appendix.

The regression tree revealed that most issues arose from non-linearity problems and the presence of a second predictor in the regression model. When the error distribution had a very small variance, all computer vision models tended to under-predict the distance. Conversely, when the error distribution had a large variance, all models tended to over-predict the distance. Therefore, Additionally, for input images representing null plots, it was expected that the models will over-predict the distance, as explained in Section \ref{null-distribution-of-the-approximated-distance}.

@@

model_pred %>%
  ggplot() +
  geom_hline(yintercept = 0, alpha = 0.5) +
  # geom_point(aes(effect_size, effect_size - vss), alpha = 0.03) +
  annotate("rect", 
           ymin = min(model_pred$effect_size - model_pred$vss),
                ymax = 0,
                xmin = min(model_pred$vss),
                xmax = max(model_pred$vss),
            fill = "#66c2a5",
            alpha = 0.3) +
  geom_point(data = NULL, aes(5, 0, col = "over-prediction"), shape = 15) +
  annotate("rect", 
           ymin = 0,
                ymax = max(model_pred$effect_size - model_pred$vss),
                xmin = min(model_pred$vss),
                xmax = max(model_pred$vss),
            fill = "#fc8d62",
            alpha = 0.3) +
  geom_point(data = NULL, aes(5, 0, col = "under-prediction"), shape = 15) +
  geom_hex(aes(vss, effect_size - vss), bins = 25) +
  geom_smooth(aes(vss, effect_size - vss), se = FALSE, col = "#e34a33") +
  facet_grid(type ~ res) +
  coord_fixed() +
  ylab(expression(D - hat(D))) +
  xlab(expression(hat(D))) +
  theme_light() +
  scale_fill_continuous(limits = c(1, 200), trans = "log10", low = "#56B1F7", high = "#132B43") +
  scale_color_manual(values = c("under-prediction" = scales::alpha("#fc8d62", 0.3),
                                "over-prediction" = scales::alpha("#66c2a5", 0.3)),
                     breaks = c("under-prediction", "over-prediction")) +
  guides(col = guide_legend(title = NULL, override.aes = list(size = 8), order = 1))
```

@@

## Model Violations Index (MVI)

Figures \ref{fig:poly-index} and \ref{fig:heter-index} display the residual plots for fitted models exhibiting varying degrees of non-linearity and heteroskedasticity. Each residual plot's Model Violation Index (MVI) is computed using Equation \ref{eq:mvi} with $C = 10$. When $\text{MVI} > 8$, the visual patterns are notably strong and easily discernible by humans. In the range $6 < \text{MVI} < 8$, the visibility of the visual pattern diminishes as MVI decreases. Conversely, when $\text{MVI} < 6$, the visual pattern tends to become relatively faint and challenging to observe. Table \ref{tab:mvi} provides a summary of the MVI usage and it is applicable to other linear regression models.

```{r mvi}
data.frame(degree = c("Strong", "Moderate", "Weak"),
           range = c("$\\text{MVI} > 8$", "$6 < \\text{MVI} < 8$", "$\\text{MVI} < 6$")) %>%
  kableExtra::kable("latex", 
                    booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Degree of model violations", "Range ($C$ = 10)"),
                    caption = "Degree of model violations or the strength of the visual signals according to the Model Violation Index (MVI). The constant $C$ is set to be 10.")
```

```{r poly-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of non-linearity violations. The Model violations index (MVI) is displayed atop each residual plot. The non-linearity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$."}
set.seed(1495)
if (!file.exists(here::here("paper/data/poly_index.rds"))) {
  e_sigma_factor <- 3
  keras_mod <- keras::load_model_tf(here::here("paper/data/phn_v2_32.keras"))
  poly_dgp <- phn_model(j = 3, sigma = 0.1^e_sigma_factor)
  ori_dat <- poly_dgp$gen(300)

  vss <- c()
  actual_ss <- c()
  dat_combined <- tibble()

  for (e_sigma in seq(0.7, 1.8, length.out = 20)^e_sigma_factor) {
    poly_dgp <- phn_model(j = 3, sigma = e_sigma)
    dat <- poly_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e, k = ori_dat$k * e_sigma / 0.1^e_sigma_factor))
    mod <- lm(y ~ x1, data = dat)
    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
    vss <- c(vss, my_vi$vss())
    actual_ss <- c(actual_ss, log(poly_dgp$sample_effect_size(dat) + 1))

    dat_combined <- bind_rows(dat_combined, mutate(dat, e_sigma = e_sigma))
  }

  saveRDS(dat_combined, here::here("paper/data/poly_index.rds"))
  saveRDS(vss, here::here("paper/data/poly_index_vss.rds"))
}

dat_combined <- readRDS(here::here("paper/data/poly_index.rds"))
vss <- readRDS(here::here("paper/data/poly_index_vss.rds"))

dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~e_sigma, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10))
```

```{r heter-index, fig.pos = "!h", fig.cap = "Residual plots generated from fitted models exhibiting varying degrees of heteroskedasticity violations. The Model violations index (MVI) is displayed atop each residual plot. The heteroskedasticity patterns are relatively strong for $MVI > 8$, and relatively weak for $MVI < 6$."}
set.seed(1495)
if (!file.exists(here::here("paper/data/heter_index.rds"))) {
  b_factor <- 2
  keras_mod <- keras::load_model_tf(here::here("paper/data/phn_v2_32.keras"))
  heter_dgp <- phn_model(include_z = FALSE, b = 0)
  ori_dat <- poly_dgp$gen(300)

  vss <- c()
  actual_ss <- c()
  dat_combined <- tibble()

  for (b in rev(seq(0.2, 4, length.out = 20))^b_factor) {
    heter_dgp <- phn_model(include_z = FALSE, b = b)
    dat <- heter_dgp$gen(300, computed = list(x1 = ori_dat$x1, e = ori_dat$e))
    mod <- lm(y ~ x1, data = dat)
    my_vi <- autovi::auto_vi(fitted_mod = mod, keras_mod = keras_mod)
    vss <- c(vss, my_vi$vss())
    actual_ss <- c(actual_ss, log(heter_dgp$sample_effect_size(dat) + 1))

    dat_combined <- bind_rows(dat_combined, mutate(dat, nb = -b))
  }

  saveRDS(dat_combined, here::here("paper/data/heter_index.rds"))
  saveRDS(vss, here::here("paper/data/heter_index_vss.rds"))
}

dat_combined <- readRDS(here::here("paper/data/heter_index.rds"))
vss <- readRDS(here::here("paper/data/heter_index_vss.rds"))

dat_combined %>%
  autovi::AUTO_VI$plot_resid() +
  facet_wrap(~nb, scales = "free", 
             labeller = function(...){ 
               tibble(nb = map(glue("'MVI' == {format(vss - log(300) + 10, digits = 3)}"), str2expression))
               },
             ncol = 5) +
  theme(strip.text = element_text(size = 10))
```

## Comparison with human visual inference and conventional tests

@@

For each lineup used in @li2023plot, there is one actual residual plot and 19 null plots. While the distance $D$ for the actual residual plot depends on the underlying data generating process, the distance $D$ for the null plots is zero. We have used our optimized computer vision model to predict $\hat{D}$ for both the actual residual plots and the null plots. Additionally, the appropriate conventional tests including the Ramsey Regression Equation Specification Error Test (RESET) [@ramsey1969tests] for non-linearity and the Breusch-Pagan test  [@breusch1979simple] for heteroskedasticity were applied on the same data for comparison.

@@

Likewise, Figure \ref{fig:conv-mosaic} and Table \ref{tab:conv-table} provide a summary of the agreement between decisions made by the computer vision model and conventional tests. The agreement rates between conventional tests and the computer vision model are 65.37% and 57.29% for residual plots containing heteroskedasticity and non-linearity patterns, respectively. These figures are lower than those presented in Table \ref{tab:human-table}, indicating that the computer vision model exhibits behaviour more akin to visual tests than conventional tests. Additionally, conventional tests tend to reject residual plots almost exclusively when the computer vision model does, but they also reject plots when the computer vision model does not. This suggests that conventional tests are more sensitive than the computer vision model.

When plotting the decision against the distance, as illustrated in Figure \ref{fig:power}, several observations emerge. Firstly, for the computer vision model, when the distance $D > 3$, almost all residual plots are rejected. However, for visual tests conducted by humans, the threshold is higher, set at $D > 5$. And conventional tests exhibit some non-rejection cases for $D > 3$. Moreover, the computer vision model tends to have fewer rejected cases than conventional tests when $D < 2$. However, visual tests demonstrate the lowest sensitivity to residual plots with small distances. Therefore, tests based on the computer vision model are less sensitive to small deviations from model assumptions than conventional tests but more sensitive to moderate deviations. Rejection decisions are fitted by logistic regression models with no intercept terms but an offset equals to $\text{log}(0.05/0/95)$. The fitting curves for the computer vision model fall between those of conventional tests and visual tests for non-linearity and are slightly below conventional tests for heteroskedasticity. There is still potential to refine the computer vision model to better align its behaviour with visual tests.

@@

Figure \ref{fig:delta} indicates that the weighted detection rate increases as the $\delta$-difference increases, particularly when the $\delta$-difference is greater than zero. A negative $\delta$-difference suggests that there is at least one null plot in the lineup with a stronger visual signal than the actual residual plot. In some instances, the weighted detection rate is close to one, yet the $\delta$-difference is negative. This discrepancy implies that the distance measure, or the approximated distance, may not perfectly reflect actual human behaviour.

@@

```{r human-table}
experiment %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  group_by(type) %>%
  summarise(n = format(n()), agree = format(sum(model_reject == reject)), rate = format(mean(model_reject == reject), digits = 4)) %>%
  kableExtra::kable("latex", 
                    booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Violations", "\\#Samples", "\\#Agreements", "Agreement rate"),
                    caption = "Summary of decisions made by visual tests conducted by human and decisions made by the computer vision model.")
```

```{r conv-table}
experiment %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  group_by(type) %>%
  summarise(n = format(n()), agree = format(sum(model_reject == conventional_reject)), rate = format(mean(conventional_reject == reject), digits = 4)) %>%
  kableExtra::kable("latex", 
                    booktabs = TRUE,
                    escape = FALSE,
                    col.names = c("Violations", "\\#Samples", "\\#Agreements", "Agreement rate"),
                    caption = "Summary of decisions made by conventional tests and decisions made by the computer vision model.")
```

@@

```{r conv-mosaic, fig.pos = "!h", fig.cap = "Rejection rate ($p$-value $\\leq0.05$) of conventional test conditional on the computer vision model decision on non-linearity (left) and heteroskedasticity (right) lineups displayed using a mosaic plot. The conventional test rejects more frequently than the computer vision model, and (almost) only rejects when the computer vision model does."}
library(ggmosaic)

experiment %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  mutate(type = ifelse(type == "polynomial", "non-linearity", type)) %>%
  mutate(type = ifelse(type == "non-linearity", "Non-linearity", "Heteroskedasticity")) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity"))) %>%
  mutate(model_reject = ifelse(model_reject, "Reject", "No")) %>%
  mutate(conventional_reject = ifelse(conventional_reject, "Reject", "No")) %>%
  mutate(across(c(model_reject, conventional_reject), ~factor(.x, levels = c("Reject", "No")))) %>%
  ggplot() +
  geom_mosaic(aes(x = ggmosaic::product(conventional_reject, model_reject), 
                  fill = conventional_reject)) +
  facet_grid(~type) +
  ylab("Conventional tests reject") +
  xlab("Computer vision model rejects") +
  # labs(fill = "Visual tests reject") +
  scale_fill_brewer("", palette = "Dark2") +
  theme_bw() +
  theme(legend.position = "none") +
  coord_fixed()
```

```{r power, fig.pos = "!h", fig.cap = "Comparison of decisions made by visual tests, conventional tests and the computer vision model. The data points are fitted with logistic regression models with no intercept but an offset equals to $\\text{log}(0.05/0.95)$."}

max_actual_ss <- max(filter(experiment, !attention_check, !null_lineup)$actual_ss)
max_poly_ss <- max(filter(experiment, !attention_check, !null_lineup, type == "polynomial")$actual_ss)
max_heter_ss <- max(filter(experiment, !attention_check, !null_lineup, type != "polynomial")$actual_ss)

model_glm_pred_poly <- glm(model_reject ~ actual_ss - 1, 
    data = filter(experiment, !attention_check, !null_lineup, type == "polynomial") %>% mutate(offset0 = log(0.05/0.95)), 
    offset = offset0,
    family = binomial()) %>%
  predict(data.frame(actual_ss = seq(0, max_poly_ss, 0.01), offset0 = log(0.05/0.95)),
          type = "response") %>%
  data.frame(d = seq(0, max_poly_ss, 0.01), type = "Non-linearity", name = "Computer vision model", value = .) %>%
  mutate(type = factor(type, levels = c("Non-linearity", "Heteroskedasticity")))
